---
title: "ST2195 Coursework"
author: "Cheng You Jun Jerral"
date: "22/2/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(ech = TRUE, message = FALSE, warning = FALSE)
```

## Load in the libaries and the dataset

Combine the yearly data from 2000-2002, and inner join plane_data to the combined yearly data

```{r}
library(tidyverse)
library(ggplot2)
library(scales)
library(lubridate)

#Reading csv files
carriers <- read.csv("carriers.csv", header = TRUE, sep = ",")
plane_data <- read.csv("plane-data.csv", header = TRUE, sep = ",")
airports <- read.csv("airports.csv", header = TRUE, sep = ",")
year_2000 <- read.csv("2000.csv", header = TRUE, sep = ",")
year_2001 <- read.csv("2001.csv", header = TRUE, sep = ",")
year_2002 <- read.csv("2002.csv", header = TRUE, sep = ",")

#Combine all years together into one big df called df_combined
df_combined <- rbind(year_2000, year_2001, year_2002)

#Cleaning empty data in plane_data
plane_data <- plane_data[-which(plane_data$type==""),]
plane_data <- filter(plane_data, year != "None")
plane_data <- filter(plane_data, year <= "2002")

#Changing to date format for filtering after this
plane_data$issue_date <- as.Date(plane_data$issue_date, "%m/%d/%Y")

#Filtering issue dates that are before 01 Jan 2003
plane_data <- plane_data %>%
  select(tailnum, type, manufacturer, issue_date, model, status, aircraft_type, engine_type, year) %>%
  filter(issue_date <= as.Date("2003-01-01"))
plane_data$year <- as.numeric(plane_data$year)
plane_data <- rename(plane_data, year_built = year)

#Re-indexing the row numbers
rownames(plane_data) <- 1:nrow(plane_data)

#Renaming columns to inner join
plane_data <- rename(plane_data, TailNum = tailnum)
carriers <- rename(carriers, UniqueCarrier = Code)

#Inner join plane_data into df_combined
df_combined <- df_combined %>% inner_join(plane_data)
df_combined <- df_combined %>% inner_join(carriers)
```

## Question 1: What is the best time of day, day of the week, and time of year to fly to minimise delays?

```{r}
#Create new data frame, df_best
df_best <- df_combined[, c("Year", "Month", "DayofMonth", "DayOfWeek", "DepTime", "ArrTime", "ArrDelay", "DepDelay")]

#Remove NA values from data frame
df_best <- na.omit(df_best)

#Changing from numeric to time format
df_best$DepTime <- substr(as.POSIXct(sprintf("%04.0f", df_best$DepTime), format='%H%M'), 12, 16)
df_best$ArrTime <- substr(as.POSIXct(sprintf("%04.0f", df_best$ArrTime), format='%H%M'), 12, 16)

#Get day name from day of week
#Using lubridate to use wday
df_best$DayOfWeek <- wday(df_best$DayOfWeek, week_start = 1, label=TRUE, abbr=FALSE)

#Create new column DelayStatus, TotalDelay, Time Classification and Season
df_best <- df_best %>%
  add_column(DelayStatus = "", .after = "DepDelay")

df_best <- df_best %>%
  add_column(Season = "", .after = "DayOfWeek")

df_best <- df_best %>%
  add_column(TimeClass = "", .after = "DayOfWeek")

#Factoring in the 15 min grace period, assuming that it is already taken into account in the data
df_best <- df_best %>%
  mutate(DelayStatus = case_when(DepDelay >= 15 ~ "Delayed",
                                 DepDelay == 0 ~ "On Time",
                                 TRUE ~ "Early"))

df_best$DelayStatus <- as.factor(df_best$DelayStatus)
df_best$DelayStatus <- factor(df_best$DelayStatus, levels = c("Early", "On Time", "Delayed"))
df_best <- df_best %>%
  mutate(Season = case_when(Month %in% 3:5 ~ "Spring",
                            Month %in% 6:8 ~ "Summer",
                            Month  %in%  9:11 ~ "Autumn",
                            TRUE ~ "Winter"))

df_best$Season <- as.factor(df_best$Season)
df_best$Season <- factor(df_best$Season, levels = c("Spring", "Summer", "Autumn", "Winter"))

#Using DepTime as the reference for the time period
df_best <- df_best %>%
  mutate(TimeClass = case_when(DepTime >= "06:00" & DepTime <= "11:59" ~ "Morning",
                               DepTime >= "12:00" & DepTime <= "16:59" ~ "Afternoon",
                               DepTime >= "17:00" & DepTime <= "19:59" ~ "Evening",
                               TRUE ~ "Night"))

df_best$TimeClass <- as.factor(df_best$TimeClass)
df_best$TimeClass <- factor(df_best$TimeClass, levels = c("Morning", "Afternoon", "Evening", "Night"))

#Plots
#Using ggplot2 and scales
ggplot(df_best, aes(x = DelayStatus, fill = TimeClass)) + 
  geom_bar(position = "dodge", stat="count") +
  geom_label(stat = "count", aes(label = ..count..), position= position_dodge(width=0.9), vjust = -0.25, show.legend = FALSE) +
  ggtitle("Delay Status by Time of Day") + 
  xlab("Delay Status") + ylab("Count") + 
  labs(fill = "Time Of Day") + 
  scale_y_continuous(labels = comma)
  
ggplot(df_best, aes(x = DelayStatus, fill = DayOfWeek)) + 
  geom_bar(position = "dodge", stat="count") +
  geom_label(stat = "count", aes(label = ..count..), position= position_dodge(width=0.9), vjust = -0.25, show.legend = FALSE) +
  ggtitle("Delay Status by Day Of Week") + 
  xlab("Delay Status") + ylab("Count") + 
  labs(fill = "Day of Week") + 
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("#f54287",
                               "#8fce00",
                               "#f1c232",
                               "#9fc5e8",
                               "#d49fe8",
                               "#d3ffce",
                               "#3399ff"))

ggplot(df_best, aes(x = DelayStatus, fill = Season)) + 
  geom_bar(position = "dodge", stat="count") +
  geom_label(stat = "count", aes(label = ..count..), position= position_dodge(width=0.9), vjust = -0.25, show.legend = FALSE) +
  ggtitle("Delay Status by Season") +
  xlab("Delay Status") + ylab("Count") +
  labs(fill = "Season") +
  scale_y_continuous(labels = comma)
```

## Question 2: Do older planes suffer more delays?

```{r}
#Creating new column DelayStatus
df_airplaneAge <- df_combined [, c("Year", "DepDelay", "year_built")]
df_airplaneAge <- df_airplaneAge %>%
  add_column(DelayStatus = "", .after = "DepDelay")

df_airplaneAge <- df_airplaneAge %>%
  mutate(DelayStatus = case_when(DepDelay >= 15 ~ "Delayed",
                                 DepDelay == 0 ~ "On Time",
                                 TRUE ~ "Early"))
  
df_airplaneAge <- na.omit(df_airplaneAge)
df_airplaneAge$DelayStatus <- as.factor(df_airplaneAge$DelayStatus)
df_airplaneAge$DelayStatus <- factor(df_airplaneAge$DelayStatus, levels = c("Early", "On Time", "Delayed"))

#Creating new column, Plane_Age and finding the age of plane
df_airplaneAge <- df_airplaneAge %>%
  add_column(Plane_Age = "", .after = "year_built")
  
df_airplaneAge$Plane_Age <- df_airplaneAge$Year - df_airplaneAge$year_built
  
df_airplaneAge <- df_airplaneAge %>%
  mutate(Plane_Age_Range = case_when(Plane_Age %in% 0:10 ~ "0-10",
                                     Plane_Age %in% 11:20 ~ "11-20",
                                     TRUE ~ "20+"))
df_airplaneAge$Plane_Age_Range <- as.factor(df_airplaneAge$Plane_Age_Range)
df_airplaneAge$Plane_Age_Range <- factor(df_airplaneAge$Plane_Age_Range, levels = c("0-10", "11-20", "20+"))

#Plotting of graph
ggplot(df_airplaneAge, aes(x = Plane_Age_Range, fill = DelayStatus)) + 
  geom_bar(position = "dodge", stat="count") +
  geom_label(stat = "count", aes(label = ..count..), position= position_dodge(width=0.9), vjust = -0.25, show.legend = FALSE) +
  ggtitle("Plane Age against Delay Status") + 
  xlab("Plane Age Range") + ylab("Count") + 
  labs(fill = "Delay Status") + 
  scale_y_continuous(labels = comma)
```

## Question 3: How does the number of people flying between different locations change over time?

```{r}
#Create new dataframe, df_people
df_people <- df_combined [, c("Year", "Month", "Origin", "Dest")]
airports <- rename(airports, Origin = iata)
df_people <- df_people %>% inner_join(airports)
df_people <- rename(df_people, OriAirport = airport)

df_people <- df_people %>%
  mutate(Month = case_when(Month == 1 ~ "Jan",
                           Month == 2 ~ "Feb",
                           Month == 3 ~ "Mar",
                           Month == 4 ~ "Apr",
                           Month == 5 ~ "May",
                           Month == 6 ~ "Jun",
                           Month == 7 ~ "Jul",
                           Month == 8 ~ "Aug",
                           Month == 9 ~ "Sep",
                           Month == 10 ~ "Oct",
                           Month == 11 ~ "Nov",
                           TRUE ~ "Dec"))

df_people$Month <- factor(df_people$Month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

#Count the number of flights from each Origin
flight_count <- df_people %>%
  count(Origin) %>%
  arrange(desc(n))
flight_count

#Choose ATL(Atlanta) as the origin, since it has the highest flight count, and filtering the other Origins out
df_people <- filter(df_people, Origin == "ATL")

#Count the number of flights from ATL(Atlanta) to each Destination
dest_count <- df_people %>%
  count(Dest) %>%
  arrange(desc(n)) %>%
  slice(1:3)
dest_count

#Filter out the other Origins except EWR, LGA and ORD
remove_dest <- c("EWR", "LGA", "ORD")
df_people <- filter(df_people, Dest %in% remove_dest)

#Plotting of bar plot
ggplot(df_people, aes(x = Month, fill = Dest)) + 
  geom_bar(position = "dodge", stat="count") +
  geom_label(stat = "count", aes(label = ..count..), position= position_dodge(width=0.9), vjust = -0.25, show.legend = FALSE) +
  ggtitle("Number of Flights per Month from ATL(Atlanta)") + 
  xlab("Month") + ylab("Number Of Flights") + 
  labs(fill = "Destination") + 
  scale_y_continuous(labels = comma)
```

## Question 4: Can you detect cascading failures as delays in one airport create delays in others?

```{r}
#Create new data frame, df_cascading_dest
df_cascading_dest <- df_combined[, c("Year", "Month", "DayofMonth", "FlightNum", "TailNum", "CRSDepTime", "DepTime", "DepDelay", "CRSArrTime", "ArrTime", "ArrDelay", "Origin", "Dest")]
df_cascading_dest <- filter(df_cascading_dest, DepDelay >= 15)
df_cascading_dest <- filter(df_cascading_dest, ArrDelay > 0)

#Count the number of flights to each Destination(ATL)
cascading_dest_count <- df_cascading_dest %>%
  count(Dest) %>%
  arrange(desc(n))

#Choose ATL(Atlanta) as the destination, since it has the highest flight delay count, and filtering the other destinations out
df_cascading_dest <- filter(df_cascading_dest, Dest == "ATL")


##Create new data frame, df_cascading_origin
df_cascading_origin <- df_combined[, c("Year", "Month", "DayofMonth", "FlightNum", "TailNum", "CRSDepTime", "DepTime", "DepDelay", "CRSArrTime", "ArrTime", "ArrDelay", "Origin", "Dest")]
df_cascading_origin <- filter(df_cascading_origin, DepDelay >= 15)
df_cascading_origin <- filter(df_cascading_origin, ArrDelay > 0)


#Choose ATL(Atlanta) as the origin, since it has the highest flight delay count, and filtering the other Origin out
df_cascading_origin <- filter(df_cascading_origin, Origin == "ATL")

#Full Join the 2 tables
cascading_main <- df_cascading_dest %>% full_join(df_cascading_origin)
cascading_main <- arrange(cascading_main, Year, Month, DayofMonth, TailNum)
cascading_main <- cascading_main %>% 
  group_by(TailNum)

#Changing from numeric to time format
cascading_main$DepTime <- substr(as.POSIXct(sprintf("%04.0f", cascading_main$DepTime), format='%H%M'), 12, 16)
cascading_main$ArrTime <- substr(as.POSIXct(sprintf("%04.0f", cascading_main$ArrTime), format='%H%M'), 12, 16)
cascading_main$CRSDepTime <- substr(as.POSIXct(sprintf("%04.0f", cascading_main$CRSDepTime), format='%H%M'), 12, 16)
cascading_main$CRSArrTime <- substr(as.POSIXct(sprintf("%04.0f", cascading_main$CRSArrTime), format='%H%M'), 12, 16)
cascading_main
```

## Question 5: Use the available variables to contrust a model that predicts delays

```{r}
#Load data
#Select 50 flights at random
library(caTools)
set.seed(1)
ml_delay <- data.frame(df_combined)
ml_delay<- select(ml_delay, -CancellationCode, -CarrierDelay, -WeatherDelay, -NASDelay, -SecurityDelay, -LateAircraftDelay, -type, -manufacturer, -issue_date, -model, -status, -aircraft_type, -engine_type, -year_built, -Description)
  ml_delay <- ml_delay %>%
  filter(!is.na(DepDelay) & !is.na(ArrDelay)) %>%
  sample_n(50)

#Plotting out linear Regression
#Scatterplot
plot(x = ml_delay$DepDelay, y = ml_delay$ArrDelay, main = "Scatterplot of DepTime vs ArrTime")

#Linear Regression Plot
ggplot(data = ml_delay, mapping = aes(x = DepDelay, y = ArrDelay)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  ggtitle("Linear Regression between DepDelay and ArrDelay")
df_delayFit <- lm(ArrDelay ~ DepDelay, data = ml_delay)
summary(df_delayFit)

#Train-Test Split
#70% trainset, sample size = 50
train <- sample.split(Y = ml_delay$ArrDelay, SplitRatio = 0.7)
trainset <- subset(ml_delay, train == T)
testset <- subset(ml_delay, train == F)

#Develop model on trainset
train_model <- lm(ArrDelay ~ DepDelay + I(DepDelay^2), data = trainset)
summary(train_model)

#Apply train_model from trainset to predict on testset.
predict_test<- predict(train_model, newdata = testset)
error_test <- testset$ArrDelay - predict_test

```

